---
title: "Assignment 10: Data Scraping"
author: "Brianna Karson"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 #loading packages
library(tidyverse)
library(rvest)
getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 #setting scraping website
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
#water system name
water_system <- webpage %>% 
html_nodes("div+ table tr:nth-child(1) td:nth-child(2)")%>% 
    html_text()
water_system

#contact person 
contact_person <- webpage %>% 
html_nodes("div+ table tr:nth-child(4) td:nth-child(2)")%>% 
    html_text()
contact_person

#ownership
ownership <- webpage %>% 
html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>% 
    html_text()
ownership

#Maximum Day Use (MGD) - for each month
max_MGD <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
 max_MGD
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 #converting scraped data into a data frame
df_1 <- data.frame(
  "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
  "Year" = rep(2024,12),
  "Monthly_MGD" = as.numeric(max_MGD)) %>% 
  mutate(Water_System = !!water_system,
         Ownership = !!ownership,
         Contact_Person = !!contact_person,
         Date = my(paste(Month,"-",Year)))

#5 #plotting MGD over time
df_1_plot <- ggplot(df_1, 
  aes(x = Date, y = Monthly_MGD)) +
  geom_line() +
  labs(title = paste("2024 Water Usage Data for", 
       water_system), subtitle = ownership,
       y = "Maximum Day Use (MGD)",
       x = "Date") +
  scale_x_date(date_breaks = "1 month",
                date_labels = "%b") 
print(df_1_plot)
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output

```{r construct.a.scraping.function}
#6. 
#creating scraping function
scrape.it <- function(PWSID, the_year){
  
#retrieving the website contents 
website <- read_html(paste0(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
  PWSID, '&year=', the_year))

#setting elements address variables
water_system_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
contact_person_tag <- 'div+ table tr:nth-child(4) td:nth-child(2)'
ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
mgd_tag <- 'th~ td+ td'

#scraping data
the_water_system <- website %>% 
  html_nodes(water_system_tag) %>% 
  html_text()
print(the_water_system)

the_contact_person <- website %>%   
  html_nodes(contact_person_tag) %>%  
  html_text()

the_ownership <- website %>% 
  html_nodes(ownership_tag) %>% 
  html_text()

the_mgd <- website %>% 
  html_nodes(mgd_tag) %>%
  html_text()
  
#constructing data frame
data_frame <- data.frame(
  "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
  "Year" = rep(the_year, 12), #year as function input
  "Max_Daily_MGD" = as.numeric(the_mgd))  %>% 
   mutate(Water_System = !!the_water_system,
         Ownership = !!the_ownership,
         Contact_Person = !!the_contact_person,
         PWSID = !!PWSID, #PWSID as function input
         Date = my(paste(Month,"-",Year)))

#scraping etiquette
Sys.sleep(1)

#returning data frame as output
return(data_frame)
 }
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 #using the function for Durham 2020
df_2 <- scrape.it('03-32-010', 2020)

#plotting the results
df_2_plot <- ggplot(df_2, 
  aes(x = Date, y = Max_Daily_MGD)) +
  geom_line() +
  labs(title = "2020 Water Usage Data for Durham", 
      subtitle = "Municipality",
      y = "Maximum Day Use (MGD)",
      x = "Date") +
 scale_x_date(date_breaks = "1 month",
               date_labels = "%b") 

print(df_2_plot)

#showing the data frame
str(df_2) 
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 #using function for Cary 2020
df_3 <- scrape.it('03-92-020', 2020)

str(df_3) 

#combining with Durham 2020 Data
df_4 <- bind_rows(df_3, df_2)

#Plotting the comparison
durham_cary_plot <- ggplot(df_4, 
  aes(x = Date, y = Max_Daily_MGD, color = Water_System)) +
  geom_line() + 
  labs(title = "2020 Water Usage Data for Durham and Cary",
       color = "City",
       y = "Maximum Day Use (MGD)",
       x = "Date") +
  scale_x_date(date_breaks = "1 month",
                date_labels = "%b")

print(durham_cary_plot)
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
#scraping years 2018 to 2023 - Cary
the_years = rep(2018:2023)
the_PWSID = '03-92-020'

#using purrr's to apply the scrape function
the_dfs <- map(the_years, scrape.it, PWSID=the_PWSID)

#combine to single data frame
df_5 <- bind_rows(the_dfs)

#plotting
cary_years_plot <- ggplot(df_5,
       aes(x = Date, y = Max_Daily_MGD)) + 
  geom_line() + 
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Cary Water Usage Data from 2018-2024",
    y="Maximum Day Use (MGD)",
    x="Date") 

print(cary_years_plot)
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: It appears that water usage in Cary has been slightly increasing over time, however it is hard to know for sure whether the slight increase over a narrow time frame would be statistically significant.
>
